\LaTeXe\ guide for the Journal of Functional Programming

### 2019-10-20 papers/jfp/7-experience.tex
Polish S5-7.

We have described a framework that can describe, and indeed execute in
prototype form, a wide spectrum of build systems.  But our ultimate
goal is a practical one: to use these insights to construct a better
build system.  Our earlier work on the \Shake build
system~\cite{mitchell2012shake}, and applying \Shake to
building GHC~\cite{hadrian}, makes progress in that direction.

Based on the theory developed in this paper we have extended \Shake to
become \Cloud \Shake, the first cloud-capable build system to support
both early cutoff and monadic dependencies~(\S\ref{sec-implementation-cloud}),
and used it to implement GHC's (very substantial) build system, Hadrian~\cite{hadrian}.
In this section we reflect on our experience of turning theory into practice.

\subsection{Haskell as a design language}\label{sec-design-lang}

@@

It is also worth noting that we needed a rather expressive language,
to faithfully express the abstractions that seem natural in this
setting.  In the case of \hs{Task} we use a data constructor with a
polymorphic field; a higher-kinded type variable \hs{f}~\hs{::}~\hs{*}~\hs{->}~\hs{*}; an
even more abstracted kind for \hs{c}~\hs{::}~\hs{(*}~\hs{->}~\hs{*)}~\hs{->}~\hs{Constraint}; and,
of course, type classes.

Moreover, when thinking about the type constructors over which \hs{f} might
usefully range, it turned out that we could adopt the \emph{existing} abstractions
of \hs{Functor}, \hs{Applicative}, \hs{Monad} and so on.
That in turn led us to think in a new way about the taxonomy
of build systems -- see \S~\ref{sec-why-polymorphism}.

The effect of using a concrete design language went well beyond
merely \emph{expressing} our ideas: it \emph{directly influenced}
our thinking.  For example, here are definitions for a scheduler and rebuilder,
from \S~\ref{sec-concrete-impl}:
\begin{minted}[fontsize=\small,xleftmargin=5pt]{haskell}
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task@\,@(MonadState ir) k v
\end{minted}
These powerful and modular abstractions, which ultimately formed part
of the conceptual structure of the paper, emerged fairly late in the
project, as we repeatedly reviewed, redesigned, and refactored our
executable prototype.  It is hard to believe that we could have
developed them without the support of Haskell as a design language.

@@

The original design of \Shake has not changed since the initial paper, but the
implementation has continued to mature -- there have been roughly 5,000
subsequent commits to the \Shake
project\footnote{\url{https://github.com/ndmitchell/shake}}. These commits add
concepts like \emph{resources} (for handling situations when two build tasks
contend on a single external resource), rewriting serialisation to be faster,
documentation including a website\footnote{\url{https://shakebuild.com}}, and
add a lot of tests. The biggest change in that time period was an implementation
change: moving from
blocking threads to continuations for the suspending scheduler.
But otherwise, almost all external and internal details remain the same\footnote{
  The most visible change is purely notational: switching from \hs{*>} to \hs{@\%@>}
  for defining rules, because a conflicting \hs{*>} operator was added to the
  Haskell \hs{Prelude}.}.
We consider the lack of change suggestive that \Shake is based on fundamental
principles -- principles we can now name and describe as a consequence
of this paper.

There are two main
aspects to the original \Shake paper~\cite{mitchell2012shake} that are described
more clearly in this paper. Firstly, the rebuilder can now be described using
verifying step traces~\S\ref{sec-step-traces}, with a much clearer relationship
to the unoptimised verifying traces of~\S\ref{sec-verifying-traces}. Secondly,
in the original paper the tasks (there called ``actions'')
were described in continuation-passing style using the data type:\footnote{
The original paper uses concrete types \hs{Key} and \hs{Value}. Here we
generalise these types to \hs{k} and \hs{v}, and also add \hs{a} so that
\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}}
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Action k v a = Finished a
                  | Depends k (v -> Action k v a)
\end{minted}
\vspace{1mm}

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
instance Monad (Action k v) where
    return = Finished
    Finished x    >>= f = f x
    Depends ds op >>= f = Depends ds $ \v -> op v >>= f

toAction :: Task Monad k v -> Action k v v
toAction (Task run) = run $ \k -> Depends k Finished

fromAction :: Action k v v -> Task Monad k v
fromAction x = Task $ \fetch -> f fetch x
  where
    f _     (Finished v  ) = return v
    f fetch (Depends d op) = fetch d >>= f fetch . op
\end{minted}
\vspace{1mm}

@@

By describing these types
using \hs{Task} we are able to describe the differences more concisely
(\hs{Monad} vs \hs{Applicative}), use existing literature to determine what is
and isn't possible, and explore other constraints beyond just \hs{Monad} and
\hs{Applicative}.
These and other isomorphisms for \emph{second-order functionals}, i.e.
functions of the form

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
addCloud :: k -> Ver -> Ver -> [[(k, Hash v)]]
         -> v -> [FilePath] -> IO ()
lookupCloud :: (k -> m (Maybe (Hash v))) -> k -> Ver -> Ver
            -> m (Maybe (v, [[k]], IO ()))
\end{minted}
\vspace{1mm}

@@

\begin{itemize}
\item Two \hs{Ver} arguments are passed to each function. These are
      the versions of the build script, and the rule for this particular key. If
      either version changes then it is as though the key has changed, and
      nothing will match. These versions are important to avoid using stale
      build products from previous versions of the build script.
\item The list of dependencies to \hs{addCloud} is a list of lists, rather than
      a simple list. The reason is that \Shake allows a list of dependencies to
      be specified simultaneously, so they can all be built in parallel.
\item The \hs{addCloud} function also takes a list of \hs{FilePath}, being the
      files that this rule produces. These produced files include those which
      are output keys from a rule and those declared with the function
      \hs{produces}.
\item The \hs{lookupCloud} function allows an explicit \hs{Nothing} when looking
      up a dependent key, since some keys are not buildable.
\item The \hs{lookupCloud} function returns at most one result, rather than a
      list. This change was made for simplicity.
\end{itemize}

To integrate these functions into \Shake we found the most expedient route was
to leave \Shake with verifying traces, but if the verifying trace does not
match, we consult the constructive trace. By bolting constructive traces onto
the side of Shake we avoid re-engineering of the central database. We have not
found any significant downsides from the bolt-on approach thus far, so it may be
a sensible route to go even if developing from scratch -- allowing an optimised
verified trace implementation in many cases, and falling back to a more complex
implementation (requiring consulting remote servers) only rarely.

@@

While we expected the GHC build system to be the first to take advantage of
\Cloud \Shake, we were actually beaten to it by Standard Chartered who
report\footnote{\url{https://groups.google.com/d/msg/shake-build-system/NbB5kMFS34I/mZ9L4TgkBwAJ}}:

@@

To help with the final two issues -- insufficient dependencies and produced
files -- we have further enhanced the \Shake lint modes, coupling them to a
program called \cmd{FSATrace}, which detects which files are read/written by a
command line execution. Such information has been very helpful in making the GHC
build cloud ready.

@@

A small number of GHC build rules cannot be cached. These rules register new libraries
in the GHC package database, and rather than producing a output file, they mutate
a shared file. \Cloud~\Shake provides a way to manually label such
build rules to exclude them from caching.

One of the benefits of using \Shake is that we have access to high quality build
profiling information, allowing us to compute critical paths and other metrics;
see Mitchell~\shortcite{mitchell2019ghcrebuildtimes} for an overview of \Shake's
profiling features. This information has shown us, for example, that more CPUs would not help (on
unlimited CPUs the speed up would be less than 10\%), and that a handful of
build tasks (two anomalously slow Haskell compilations, and some calls to slow
single-threaded \cmd{configure}) are responsible for a significant fraction of
build time (at least 15\%).

>>>>

We have presented a framework that can describe, and indeed execute in
prototype form, a wide spectrum of build systems.  But our ultimate
goal is a practical one: to use these insights to develop better
build systems. Our earlier work on \Shake~\cite{mitchell2012shake}, and applying
\Shake to building GHC~\cite{hadrian}, makes progress in that direction.

Based on the theory developed in this paper we have extended \Shake to
become \Cloud \Shake, the first cloud-capable build system to support
both early cutoff and monadic dependencies~(\S\ref{sec-implementation-cloud}),
and used it to implement GHC's (very substantial) build system,
\Hadrian~\cite{hadrian}. In this section we reflect on our experience of turning
theory into practice.

\vspace{-1mm}
\subsection{Haskell as a Design Language}\label{sec-design-lang}

@@

It is also worth noting that we needed a rather expressive language,
to faithfully express the abstractions that seem natural in this
setting.  In the case of \hs{Task} we use a data constructor with a
polymorphic field; a higher-kinded type variable \hs{f}~\hs{::}~\hs{*}~\hs{->}~\hs{*}; an
even more abstracted kind for \hs{c}~\hs{::}~\hs{(*}~\hs{->}~\hs{*)}~\hs{->}~\hs{Constraint}; and,
of course, type classes. Our models have been translated to
Rust~\cite{translation_rust} and Kotlin~\cite{translation_kotlin}, and in both
cases there was a loss of precision due to language-specific limitations.

When thinking about the type constructors over which \hs{f} might usefully
range, it turned out that we could adopt the \emph{existing abstractions}
of \hs{Functor}, \hs{Applicative}, \hs{Monad} and so on.
That in turn led us to think in a new way about the taxonomy
of build systems -- see~\S\ref{sec-why-polymorphism}. In the other direction,
trying to express an \emph{existing build system} \Dune in our models led us to
finding a new abstraction -- the \hs{Selective} \mbox{type
class~\cite{mokhov_selective_2019}}, which turned out to be useful outside the
build systems domain.

The effect of using a concrete design language went well beyond
merely \emph{expressing} our ideas: it \emph{directly influenced}
our thinking.  For example, here are definitions for a scheduler and rebuilder,
from \S~\ref{sec-concrete-impl}:

\vspace{0.5mm}
\begin{minted}[fontsize=\small,xleftmargin=5pt]{haskell}
type Scheduler c i ir k v = Rebuilder c ir k v -> Build c i k v
type Rebuilder c   ir k v = k -> v -> Task c k v -> Task@\,@(MonadState ir) k v
\end{minted}
\vspace{0.5mm}

These powerful and modular abstractions, which ultimately formed part
of the conceptual structure of the paper, emerged fairly late in the
project, as we repeatedly reviewed, redesigned, and refactored our
executable prototypes. It is hard to believe that we could have
developed them without the support of Haskell as a design language.

@@

The original design of \Shake has not changed since the initial paper, but the
implementation has continued to mature -- there have been roughly 5,000
subsequent commits to the \Shake
project\footnote{See \url{https://github.com/ndmitchell/shake}.}. These commits
add concepts like \emph{resources} (for handling situations when two build tasks
contend on a single external resource), rewriting serialisation to be faster,
documentation including a website\footnote{See \url{https://shakebuild.com}.},
and add a lot of tests. The biggest change in that time period was an
implementation change: moving from
blocking threads to continuations for the suspending scheduler.
But otherwise, almost all external and internal details remain the
  same\footnote{The most visible change is purely notational: switching from
  \hs{*>} to \hs{@\%@>} for defining rules, because a conflicting \hs{*>}
  operator was added to the Haskell \hs{Prelude}.}.
We consider the lack of change suggestive that \Shake is based on fundamental
principles -- principles we can now name and describe as a consequence
of this paper.

There are two main
aspects to the original \Shake paper~\cite{mitchell2012shake} that are described
more clearly in this paper. Firstly, the rebuilder can now be described using
verifying step traces~\S\ref{sec-step-traces}, with a much clearer relationship
to the unoptimised verifying traces of~\S\ref{sec-verifying-traces}. Secondly,
in the original paper the tasks (there called ``actions'')
were described in continuation-passing style using the data
type\footnote{The original paper uses concrete types \hs{Key} and \hs{Value}.
Here we generalise these types to~\hs{k} and~\hs{v}, and also add~\hs{a} so that
\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}.}:
\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Action k v a = Finished a
                  | Depends k (v -> Action k v a)
\end{minted}
\vspace{0.5mm}

@@

\vspace{0.5mm}
\begin{minted}[xleftmargin=10pt]{haskell}
instance Monad (Action k v) where
    return = Finished
    Finished x    >>= f = f x
    Depends ds op >>= f = Depends ds $ \v -> op v >>= f
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
toAction :: Task Monad k v -> Action k v v
toAction (Task run) = run (\k -> Depends k Finished)
\end{minted}
\begin{minted}[xleftmargin=10pt]{haskell}
fromAction :: Action k v v -> Task Monad k v
fromAction x = Task (\fetch -> f fetch x)
  where
    f _     (Finished v  ) = return v
    f fetch (Depends d op) = fetch d >>= f fetch . op
\end{minted}
\vspace{0.5mm}

@@

By expressing these types
using \hs{Task} we are able to describe the differences more concisely
(\hs{Monad} vs \hs{Applicative}), use existing literature to determine what is
and isn't possible, and explore other constraints beyond just \hs{Monad} and
\hs{Applicative}.
These and other isomorphisms for \emph{second-order functionals}, i.e.
functions of the form

@@

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
addCloud :: k -> Ver -> Ver -> [[(k, Hash v)]] -> v -> [k] -> IO ()
lookupCloud :: (k -> m (Maybe (Hash v))) -> k -> Ver -> Ver
            -> m (Maybe (v, [[k]], IO ()))
\end{minted}
\vspace{1mm}

@@

\begin{itemize}
\item Two \hs{Ver} arguments are passed to each function. These are
      the versions of the build script, and the rule for this particular key. If
      either version changes then it is as though the key has changed, and
      nothing will match. These versions are important to avoid using stale
      build products from previous versions of the build script.
\item The list of dependencies to \hs{addCloud} is a list of lists, rather than
      a simple list. The reason is that \Shake allows a list of dependencies to
      be specified simultaneously, so they can all be built in parallel.
\item The \hs{addCloud} function also takes a list of keys \hs{[k]}, being the
      files that this rule produces. These produced files include those which
      are output keys from a rule and those declared with the function
      \hs{produces}.
\item The \hs{lookupCloud} function allows an explicit \hs{Nothing} when looking
      up a dependent key, since some keys are not buildable.
\item The \hs{lookupCloud} function returns at most one result, rather than a
      list. This change was made for simplicity.
\end{itemize}

To integrate these functions into \Shake we found the most expedient route was
to leave \Shake with verifying traces, but if the verifying trace does not
match, we consult the constructive trace. By bolting constructive traces onto
the side of \Shake we avoid re-engineering of the central database. We have not
found any significant downsides from the bolt-on approach thus far, so it may be
a sensible route to go even if developing from scratch -- allowing an optimised
verified trace implementation in many cases, and falling back to a more complex
implementation (requiring consulting remote servers) only rarely.

@@

While we expected the GHC build system to be the first to take advantage of
\Cloud \Shake, we were actually beaten to it by Standard Chartered who
report\footnote{\mbox{\url{https://groups.google.com/d/msg/shake-build-system/NbB5kMFS34I/mZ9L4TgkBwAJ}}}:

@@

To help with the final two issues -- insufficient dependencies and produced
files -- we have further enhanced the \Shake lint modes, coupling them to a
program called \cmd{FSATrace}, which detects which files are read/written by a
command line execution. Such information has been very helpful in making the GHC
build cloud ready~\cite{cloud_hadrian}.

@@

A small number of GHC build rules cannot be cached. These rules register
libraries in the package database, and rather than producing files, they
mutate a shared file. \Cloud~\Shake provides a way to manually label such
build rules to exclude them from caching.

One of the benefits of using \Shake is that we have access to high quality build
profiling information, allowing us to compute critical paths and other metrics;
see Mitchell~\shortcite{mitchell2019ghcrebuildtimes} for an overview of \Shake's
profiling features. This information has shown us, for example, that more CPUs would not help (on
unlimited CPUs the speed up would be less than 10\%), and that a handful of
build tasks (two anomalously slow Haskell compilations, and calls to slow
single-threaded \cmd{configure}) take up a significant fraction of build time
(at least 15\%).
